\section{Itanium}
\label{itanium}
\myindex{Itanium}

Anche se fallita, Intel Itanium (\ac{IA64}) è un'architettura molto interessante.

Mentre le \ac{OOE} CPU decidono come riordinare le istruzioni ed eseguirle contemporaneamente,
\ac{EPIC} era un tentativo di spostare queste decisioni al compilatore:
per raggruppare le istruzioni alla fase di compilazione.

Questo implica a sua volta compilatori più complessi.

Ecco un esempio di codice \ac{IA64}: un semplice algoritmo criptografico dal kernel di Linux:

\lstinputlisting[caption=Linux kernel 3.2.0.4]{other/itanium/tea_from_linux.c}

Ed ecco come è stato compilato:

\lstinputlisting[caption=Linux Kernel 3.2.0.4 for Itanium 2 (McKinley)]{other/itanium/ia64_linux_3.2.0.4_mckinley.lst}

Per iniziare, tutte le istruzioni \ac{IA64} sono raggruppate in 3 gruppi.

Ogni gruppo ha una grandezza di 16 bytes (128 bits) e consiste nel modello di 3 codici (5 bits) + 3 istruzioni (41 bits ciascuno).

\IDA mostra i gruppi come 6+6+4 bytes~---dove puoi facilmente individuare il modello.

Tutte e 3 le istruzioni da ciascun gruppo, di solito si eseguono simultaneamente, finche una delle istruzioni ha un \q{stop bit}.

Supponiamo che gli ingegneri di Intel e HP raccolgono statistiche sui modelli delle istruzioni più frequenti e decidono di portare
dei tipi di modelli: a bundle code defines the instruction types in the bundle.
There are 12 of them.

For example, the zeroth bundle type is \TT{MII}, which implies 
the first instruction is Memory (load or store), the second and third ones are I (integer instructions).

Another example is the bundle of type 0x1d: \TT{MFB}:
the first instruction is Memory (load or store), the second one is Float 
(\ac{FPU} instruction), and the third is Branch (branch instruction).

If the compiler cannot pick a suitable instruction for the relevant bundle slot, it may insert a \ac{NOP}:
you can see here the
\TT{nop.i} instructions (\ac{NOP} at the place where the integer instruction might be) or \TT{nop.m} 
(a memory instruction might be at this slot).

\ac{NOP}s are inserted automatically when one uses assembly language manually.

And that is not all. Bundles are also grouped.

Each bundle may have a \q{stop bit},
so all the consecutive bundles with a terminating bundle which has the \q{stop bit} 
can be executed simultaneously.

In practice, Itanium 2 can execute 2 bundles at once, resulting in the execution of 6 instructions at once.

So all instructions inside a bundle and a bundle group cannot interfere with each other 
(i.e., must not have data hazards).

If they do, the results are to be undefined.

Each stop bit is marked in assembly language as two semicolons (\TT{;;}) after the instruction.

So, the instructions at [90-ac] may be executed simultaneously:
they do not interfere. The next group is [b0-cc].

We also see a stop bit at 10c.
The next instruction at 110 has a stop bit too.

This implies that these instructions must be executed isolated from all others (as in \ac{CISC}).

Indeed: the next instruction at 110 uses the result from the previous one (the value in register r26),
so they cannot be executed at the same time.

Apparently, the compiler was not able to find a better way to parallelize the instructions,
in other words, to load \ac{CPU} as much as possible, hence too much stop bits and \ac{NOP}s.

Manual assembly programming is a tedious job as well: the programmer has to group the instructions manually.

The programmer is still able to add stop bits to each instructions, but this will degrade
the performance that Itanium was made for.

An interesting examples of manual \ac{IA64} assembly code can be found in the Linux kernel's sources:

\url{http://go.yurichev.com/17322}.

Another introductory paper on Itanium assembly:
[Mike Burrell, \IT{Writing Efficient Itanium 2 Assembly Code} (2010)]\footnote{\AlsoAvailableAs \url{http://yurichev.com/mirrors/RE/itanium.pdf}},
[papasutra of haquebright, \IT{WRITING SHELLCODE FOR IA-64} (2001)]\footnote{\AlsoAvailableAs \url{http://phrack.org/issues/57/5.html}}.

Another very interesting Itanium feature is the \IT{speculative execution} and the NaT (\q{not a thing}) bit,
somewhat resembling \gls{NaN} numbers: \\
\href{http://go.yurichev.com/17323}{MSDN}.

